DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-08-09 2015CompleteDerivation/completeDerivation.tex
Another pass at the kinematics text

\begin{rationale}
	The idea is that the system internal dynamics has no affect on the motion and therefore can be disregarded. Consider the motion of our cannonball under gravitational and inertial forces: its temperature or precise chemical composition does not affect the trajectory. We can assume kinematic equivalence. Consider the motion of a helicopter under gravitational and inertial forces: the pilot and the amount of fuel in the tank do affect the trajectory. We cannot assume kinematic equivalence.

	The assumption clearly cannot be taken in general: kinematic variables may not fully identify the particle state of a material. Yet, the assumption is fundamental because particles of any material will \emph{at least} describe a trajectory in space and what we learn about the link between kinematic and dynamical quantities applies to all materials, even the ones where the kinematic assumption does not apply.

	The first natural question to ask is: why does state evolution always describe a trajectory in space? Because spatial extent is the one way that all materials can be decomposed. As we divide materials in smaller pieces, physical location is one of the state variable for the particles of all materials. If we assume kinematic equivalence, the kinematic variables will be enough to identify a particle state. In such a case, we will call the material \emph{minimal} to highlight the idea that the state is described by the smallest possible number of d.o.f..

	When the system is not minimal, the particles will be described by some additional independent state variables. The overall dynamics will be the spatial d.o.f. coupled with the internal d.o.f.. Therefore by studying a minimal material, we are also studying a component of the dynamics that is present in all materials.

	To be clear: we are not saying that kinematic equivalence is fundamental because everything must reduce to a material with no internal dynamics. That is: we are not saying elementary materials are minimal. While this statement seems reasonable and may fit our physical intuition, we have no a priori argument for it. We can imagine, in fact, that as we reduce the behavior of one material to its constituents we find the parts to possess richer and richer states. In a similar way that the elementary particles have electromagnetic and color charges, while their aggregation is typically neutral, we could imagine finding more and more richness in the internal dynamics. While this does not seem to happen in practice, we do not have an argument why this cannot happen.

	What we are saying is that any material is \emph{at least} a minimal material and will possess the dynamic of one coupled with the internal dynamics.

@@

As we need to describe trajectories, we introduce $\mathcal{M}$ as the $n+1$ dimensional manifold that defines space and time. A point in $\mathcal{M}$ will be identified by $x^\alpha = (t, x^i)$, its time and space coordinates. A trajectory will be given by a set of functions $x^\alpha(s)$ where $s$ is an arbitrary parametrization of the curve. What we need to understand is the link between the complete state space of the particles of a classical material and the space of all the possible trajectories under the kinematic assumption.

@@

Because of the double role of $\mathcal{M}$, we can set half of our state variables to be equal to the space-time variables: $q^\alpha = (t, q^i) = x^\alpha$. This already tells us that the trajectory is differentiable since $q^i(t)$ is differentiable. The four velocity $u^\alpha = d_s x^\alpha$ is therefore well defined and it can be used to identify trajectories. The magnitude of the four velocity can be rescaled by changing the evolution parameter, $u^\alpha = d_s x^\alpha = d_s \hat{s} \, d_{\hat{s}} x^\alpha = d_s \hat{s} \, \hat{u}^\alpha$, and is therefore not physical. Only the direction is physical so the four velocity gives us another $n$ independent variables to identify trajectories. The combined position and velocity, the kinematic variables, identify $2n + 1$ trajectories using a hyper-surface of the tangent bundle $T\mathcal{M}$. This is exactly the dimensionality of the complete state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$. We cannot add more curves, as a homeomorphism preserves dimensionality, therefore we found the suitable space for the trajectories of particles of a classical material under kinematic equivalence.

Being able to identify particles through either state or kinematic variables is not enough: we also need to be able to express the distributions of the composite states over position and velocity. As such, the differentiable distribution over state variables must become a differentiable distribution over kinematic variables. This tells us the transformation is differentiable.

@@

The metric tensor $g$ and the constant $c$ found here are indeed the ones used in the context of relativity. Yet, they have acquired a different meaning. $g$ does not measure space-time distances: it measure cardinality of possibilities, like $\omega$. Directions in space are orthogonal not because the angle between them is $90^\circ$, but because they identify independent d.o.f.. Space and time, instead, are not independent and the geometry between them is different. In other words: $g$ inherits it's signature from $\omega$.

@@

The link between $\omega$ and $g$ allows us to constrain the kind of diffeomorphism between state variables and kinematic variables. If we express the possibilities in terms of both state variables and initial conditions we have: $dq^\alpha \wedge dp_\alpha \propto dx^\alpha du_\alpha$. Since  $p_\alpha=p_\alpha(x, u)$, we have $dp_\alpha = m du_\alpha$, where $m$ is the constant of proportionality. We recognize $m$ as the inertial mass. Integrating the expression, we have $p_\alpha = m u_\alpha + A_\alpha(x)$ where $A_\alpha(x)$ is an arbitrary function. This is the most general link between velocity and conjugate momentum, which is the sum of kinetic momentum and a scalar potential.

@@

The above discussion gives the inertial mass a new physical meaning: it is the conversion constant that tells us how to convert a range of velocity to a range of conjugate momentum with the same number of possibilities. In other words: the more massive the particles, the more states per unit velocity. Intuitively, a more massive body is harder to accelerate because it has to go through more states for the same change in velocity. Note that the Hamiltonian excludes massless particles because the kinematic assumption does that implicitly. A massless particle always has the same velocity, $c$, therefore the trajectory is not enough to tell us the dynamics. The motion of the photon is not enough to tell us its energy or its momentum. Therefore it makes sense that, by assuming kinematic equivalence we derived equation valid only for massive particles.

We have seen that the inertial mass $m$, the invariant speed $c$ and the reduced Plank constant $\hbar$ all serve the similar purposes: convert one quantity to another preserving the number of possibilities. $dp=\hbar dk$ converts between ranges of wave number and conjugate momentum. $dx = c dt$ converts between ranges of time and space. $dp = m du$ converts between ranges of velocity and conjugate momentum. They allow to keep proper book keeping when mapping sets of states. Using natural units means measuring all these quantities through the cardinality of possibilities. There is a difference, though: while $c$ and $\hbar$ are simply determined by the unit system, $m$ also depends on the type of material. That is, different homogeneous materials can have different inertial mass.

@@

We should also ask whether any choice of $A$ actually represent a different case leading to different trajectories. The simplest and more insightful way to study this problem is to use the kinematic quantities $x^\alpha$ and $mu_\alpha$ as state variable. They are not going to be conjugates of each other but they do capture the entire physical description. Any change in $A$ that does not affect the relationship between $x^\alpha$ and $mu_\alpha$ does not change the physics. Expressing the Hamiltonian and the Poisson brackets in terms of the initial conditions we have:
\begin{align*}
\mathcal{H}&=\frac{1}{2m}(mu_\alpha mu^\alpha + m^2 c^2) \\
\{x^\alpha, x^\beta\} &= 0 \\
\{x^\alpha, mu_\beta\} &= \delta^\alpha_\beta \\
\{mu_\alpha, mu_\beta\} &= \partial_\alpha A_\beta - \partial_\beta A_\alpha = F_{\alpha\beta}
\end{align*}
Note how the Poisson bracket between kinetic momentum is the force tensor. The weaker the force, the more $(x^i, mu_i)$ become independent d.o.f. The force tensor, then, tells how much $(x^i, mu_i)$ fail to be independent d.o.f.. With these expressions, the commutator between any function of position and kinetic momentum can now be calculated. For example:
\begin{align*}
d_s x^\alpha &= \{x^\alpha, \mathcal{H}\} = u^\alpha \\
d_s (mu^\alpha) &= \{mu^\alpha, \mathcal{H}\} = F^{\alpha\beta} u_\beta
\end{align*}
where we recognize the covariant form of the Lorentz force.

@@

To sum up, the kinematic assumption has more inherent consequences one would expect at first. It means the particles are identified by position and velocity and therefore the system of equations is second order. It links the space of trajectories on $T\mathcal{M}$ with the space of states on $T^*\mathcal{M}$. Such link allows state space metric $\omega$ to induce the space-time metric $g$. It allows us to define Lagrangians. It constrains the motion to massive particles under scalar and vector potential forces. Moreover, it allowed to re-understand basic concepts, like the invariant speed $c$, the inertial mass $m$ and the force tensor $F_{\alpha\beta}$ in terms of state counting possibilities and d.o.f. independence.

>>>>

\begin{rationale}
	The idea is that the system internal dynamics has no affect on the motion and therefore can be disregarded. Consider the motion of our cannonball under gravitational and inertial forces: its temperature or precise chemical composition does not affect the overall trajectory. We can assume kinematic equivalence. Consider the motion of a helicopter under gravitational and inertial forces: the pilot and the amount of fuel in the tank do affect the trajectory. We cannot assume kinematic equivalence.

	Kinematic equivalence can't therefore be assumed for all systems. Yet, the assumption is fundamental because the particles of any material will \emph{at least} describe a trajectory in space and what we learn about the link between kinematic and dynamical quantities applies to all materials, even the ones where the kinematic assumption does not apply.

	The first natural question to ask is: why does state evolution always describe a trajectory in space? Because spatial extent is the one way in which all materials can be decomposed. As we divide materials in smaller pieces, physical location is one state variable we can assign to the particles of all materials. If we assume kinematic equivalence, the kinematic variables will be enough to identify a particle state. In such a case, we will call the material \emph{minimal} to highlight the idea that the state is described by the smallest possible number of d.o.f..

	When the system is not minimal, the particles will be described by some additional independent state variables. The overall dynamics will be the spatial d.o.f. coupled with the internal d.o.f.. Therefore by studying a minimal material, we are also studying a component of the dynamics that is present in all materials. In other words: any material is \emph{at least} a minimal material and will possess the dynamic needed to describe its motion in space.

@@

As we need to describe trajectories, we introduce $\mathcal{M}$ as the $n+1$ dimensional manifold that defines physical space and time. A point in $\mathcal{M}$ will be identified by $x^\alpha = (t, x^i)$, its time and space coordinates. A trajectory will be given by a set of functions $x^\alpha(s)$ where $s$ is an arbitrary parametrization of the curve. What we need to understand is the link between the complete state space of the particles of a classical material and the space of all the possible trajectories under the kinematic assumption.

@@

Because of the double role of $\mathcal{M}$, we can set half of our state variables to be equal to the space-time variables: $q^\alpha = (t, q^i) = x^\alpha$. This already tells us that the trajectory is differentiable since $q^i(t)$ is differentiable. The four velocity $u^\alpha = d_s x^\alpha$ is therefore well defined and it can be used to identify trajectories. The magnitude of the four velocity can be rescaled by changing the evolution parameter, $u^\alpha = d_s x^\alpha = d_s \hat{s} \, d_{\hat{s}} x^\alpha = d_s \hat{s} \, \hat{u}^\alpha$, and is therefore not physical. Only the direction is physical so the four velocity gives us another $n$ independent variables to identify trajectories. The combined position and velocity, the kinematic variables, identify $2n + 1$ trajectories using a hyper-surface of the tangent bundle $T\mathcal{M}$. This is exactly the dimensionality of the complete state space $\bar{\mathcal{S}} \subset T^*\mathcal{M}$. We cannot add more curves, as a homeomorphism preserves dimensionality, therefore we identified the suitable space for the trajectories of particles of a classical material under kinematic equivalence.

Being able to identify particles through either state or kinematic variables is not enough: we also need to be able to express a composite state as the distribution of particle states over position and velocity. As such, the differentiable distribution over state variables must become a differentiable distribution over kinematic variables. This tells us the transformation is differentiable.

@@

The metric tensor $g$ and the constant $c$ found here are indeed the ones used in the context of relativity. Yet, they have acquired a different meaning. $g$ does not measure space-time distances: it measure cardinality of possibilities, like $\omega$. Directions in space are orthogonal not because the angle between them is $90^\circ$, but because they identify independent d.o.f.. Space and time, instead, are not independent and the geometry between them is different. In other words: $g$ inherits it's signature from $\omega$. The geometry of space-time is the geometry of the ranges for possible states.

@@

The link between $\omega$ and $g$ allows us to constrain the kind of diffeomorphism between state variables and kinematic variables. If we express the possibilities in terms of both state variables and kinematic variables we have: $dq^\alpha \wedge dp_\alpha \propto dx^\alpha du_\alpha$. Since  $p_\alpha=p_\alpha(x, u)$, we have $dp_\alpha = m du_\alpha$, where $m$ is the constant of proportionality. We recognize $m$ as the inertial mass. Integrating the expression, we have $p_\alpha = m u_\alpha + A_\alpha(x)$ where $A_\alpha(x)$ is an arbitrary function. This is the most general link between velocity and conjugate momentum, which is the sum of kinetic momentum and a scalar potential.

@@

The above discussion gives the inertial mass a new physical meaning: it is the conversion constant that tells us how to convert a range of velocity to a range of conjugate momentum with the same number of possibilities. In other words: the more massive the particles, the more states per unit velocity. Intuitively, a more massive body is harder to accelerate because it has to go through more states for the same change in velocity. Also note that the Hamiltonian excludes massless particles because the kinematic assumption does that implicitly. A massless particle always has the same velocity, $c$, therefore the trajectory is not enough to tell us the dynamics. The motion of the photon is not enough to tell us its energy or its momentum. Therefore it makes sense that, by assuming kinematic equivalence we derived equation valid only for massive particles.

We have seen that the inertial mass $m$, the invariant speed $c$ and the reduced Plank constant $\hbar$ all serve a similar purposes: convert one quantity to another preserving the number of possibilities. $dp=\hbar dk$ converts between ranges of wave number and conjugate momentum. $dx = c dt$ converts between ranges of time and space. $dp = m du$ converts between ranges of velocity and conjugate momentum. They allow to keep proper book keeping when mapping sets of states. Using natural units means measuring all these quantities through the cardinality of possibilities. There is a difference among them, though: while $c$ and $\hbar$ are simply determined by the unit system, $m$ also depends on the type of material. That is, different homogeneous materials can have different inertial mass.

@@

We should also ask whether any choice of $A$ actually represent a different case leading to different trajectories. The simplest and more insightful way to study this problem is to use the kinematic quantities $x^\alpha$ and $mu_\alpha$ as state variable. They are not going to be conjugates of each other but they do capture the entire physical description. Any change in $A$ that does not affect the relationship between $x^\alpha$ and $mu_\alpha$ does not change the physics. Expressing the Hamiltonian and the Poisson brackets in terms of the initial conditions we have:
\begin{align*}
\mathcal{H}&=\frac{1}{2m}(mu_\alpha mu^\alpha + m^2 c^2) \\
\{x^\alpha, x^\beta\} &= 0 \\
\{x^\alpha, mu_\beta\} &= \delta^\alpha_\beta \\
\{mu_\alpha, mu_\beta\} &= \{p_\alpha, - A_\beta\} + \{-A_\alpha, p_\beta\} \\
&= \partial_\alpha A_\beta - \partial_\beta A_\alpha = F_{\alpha\beta}
\end{align*}
Note how the Poisson bracket between kinetic momentum is the force tensor. The weaker the force, the more $(x^i, mu_i)$ become independent d.o.f. The force tensor, then, tells how much $(x^i, mu_i)$ fail to be independent d.o.f.. With these expressions, the Poisson bracket between any function of position and kinetic momentum can now be calculated. For example:
\begin{align*}
d_s x^\alpha &= \{x^\alpha, \mathcal{H}\} = u^\alpha \\
d_s (mu^\alpha) &= \{mu^\alpha, \mathcal{H}\} = F^{\alpha\beta} u_\beta
\end{align*}
where we recognize the covariant form of the Lorentz force.

@@

To sum up, the kinematic assumption has more inherent consequences one would expect at first. It means the particles are identified by position and velocity and therefore the system of equations is second order. It links the space of trajectories on $T\mathcal{M}$ with the space of states on $T^*\mathcal{M}$. Such link allows state space metric $\omega$ to induce the space-time metric $g$. It allows us to define Lagrangians. It constrains the motion to massive particles under scalar and vector potential forces. Moreover, it allowed to re-understand basic concepts, like the invariant speed $c$, the inertial mass $m$ and the force tensor $F_{\alpha\beta}$ in terms of state counting and d.o.f. independence. Finally, it allowed to construct classical equivalent for anti-particles and the CPT theorem.
